---
title: 'Cervical Cancer Case Study'

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

################## MUST manually install dependencies ##################

#install.packages("caret", repos = "http://cran.r-project.org", dependencies = c("Depends", "Imports", "Suggests"))
#install.packages('e1071', dependencies=TRUE)

library(ggplot2)
library(MASS)
library(Amelia)
library(tinytex)
library(dplyr)
library(tidyverse)
library(aod)
library(pROC)
library(cowplot)
library(broom)
library(ROCR)
library(caret)
library(ggplot2)
library(ResourceSelection)
library(survival)
library(ggfortify)
library(vip)
rm(list=ls()) # Clean the environment
```

## Retrive data

```{r}

# Get original data from the same folder
cervical_cancer_data <- readxl::read_xls("cervical_cancer.xls", sheet = "Sheet1")

```

## Change columns names

```{r}

names(cervical_cancer_data)[names(cervical_cancer_data) 
    == "AGE_1"] <- "AGE"
names(cervical_cancer_data)[names(cervical_cancer_data) 
    == "CLS_1"] <- "CLS"
names(cervical_cancer_data)[names(cervical_cancer_data) 
    == "GRAD_1"] <- "GRAD"
names(cervical_cancer_data)[names(cervical_cancer_data) 
    == "CLS_1"] <- "CLS"
names(cervical_cancer_data)[names(cervical_cancer_data) 
    == "HISTOLOG_1"] <- "HISTOLOG"
names(cervical_cancer_data)[names(cervical_cancer_data) 
    == "MAXDEPTH_1"] <- "MAXDEPTH"
names(cervical_cancer_data)[names(cervical_cancer_data) 
    == "PELLYMPH_1"] <- "PELLYMPH"
names(cervical_cancer_data)[names(cervical_cancer_data) 
    == "SIZE_1"] <- "SIZE"

```

## Permanent data cleaning

```{r, results="hide"}
### Patients without a FU_DATE are excluded 
### Recurrence yes or no (using RECURRN1 NA or date)
### Recode ADJ_RAD, radiation yes, no
### Patients that died without relapse are excluded

# Patients that don't have a final check up date are excluded
cervical_cancer_data <- 
    cervical_cancer_data[complete.cases(
    cervical_cancer_data[,c("FU_DATE")]),]

# Adds new column named "RECURRN_Y_or_N"
# Column is initialized to 0 (Recurrence == No)
cervical_cancer_data$RECURRN_Y_or_N <- 0

# Set the values in RECURRN1 to 1 for Recurrence == Yes
cervical_cancer_data$RECURRN_Y_or_N[!is.na(
    cervical_cancer_data$RECURRN1)] <- 1

# Reorder new column to be right after RECURRN1 
cervical_cancer_data <- cervical_cancer_data[,c(1:12, 15, 13, 14)]

# Recode ADJ_RAD to be either radiation therapy 1 - yes, 0 - no
cervical_cancer_data$ADJ_RAD[
    cervical_cancer_data$ADJ_RAD > 0] <- 1

# Delete patients that died without relapse because they didn't have
# the full opportunity to relapse
cervical_cancer_data <- subset(cervical_cancer_data, !(RECURRN_Y_or_N==0 & DIS_STA > 0))

cervical_cancer_data
```

## Make categorical data as factor

```{r, results="hide"}

cervical_cancer_data_factor <- cervical_cancer_data

cervical_cancer_data_factor$ADJ_RAD <- 
    as.factor(cervical_cancer_data_factor$ADJ_RAD)
cervical_cancer_data_factor$CLS <- 
    as.factor(cervical_cancer_data_factor$CLS)
cervical_cancer_data_factor$GRAD <- 
    as.factor(cervical_cancer_data_factor$GRAD)
cervical_cancer_data_factor$HISTOLOG <- 
    as.factor(cervical_cancer_data_factor$HISTOLOG)
cervical_cancer_data_factor$MARGINS <- 
    as.factor(cervical_cancer_data_factor$MARGINS)
cervical_cancer_data_factor$PELLYMPH <- 
    as.factor(cervical_cancer_data_factor$PELLYMPH)
cervical_cancer_data_factor$DIS_STA <- 
    as.factor(cervical_cancer_data_factor$DIS_STA)
# cervical_cancer_data_factor$RECURRN_Y_or_N <- 
#    as.factor(cervical_cancer_data_factor$RECURRN_Y_or_N)
# The hoslem.test doesn't work if the response is a factor

cervical_cancer_data_factor

```

## Observe data

```{r}

missmap(subset(cervical_cancer_data_factor, select = -RECURRN1), 
    main = "Missing values vs observed")

```


## Data cleaning based on model

```{r, results="hide"}
### Omit NA rows based on selected columns
### Start with cleaning for full model

cervical_cancer_data_clean <- cervical_cancer_data_factor

# Include column names to have NA rows deleted
cervical_cancer_data_clean <- 
    cervical_cancer_data_clean[complete.cases(
    cervical_cancer_data_clean[,c("ADJ_RAD",	"AGE",	"CLS", 
    "DIS_STA", "GRAD", "HISTOLOG", "MARGINS",	"MAXDEPTH",
    "PELLYMPH", "RECURRN_Y_or_N", "SIZE")]),]

# Delete GRAD missing values (Delete rows with a value of 
# 0 – indicating a missing value.)
cervical_cancer_data_clean <- subset(cervical_cancer_data_clean,
    !(cervical_cancer_data_clean$GRAD==0))

# Only NA's left in column "RECURRN1" as NA in this column means Recurrence == No

cervical_cancer_data_clean

```

## Explore data

```{r, include=FALSE}

# Relapse rate by ADJ_RAD
plot(prop.table(table(cervical_cancer_data_clean$ADJ_RAD, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by CLS
plot(prop.table(table(cervical_cancer_data_clean$CLS, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by DIS_STA
plot(prop.table(table(cervical_cancer_data_clean$DIS_STA, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by GRAD
plot(prop.table(table(cervical_cancer_data_clean$GRAD, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by HISTOLOG
plot(prop.table(table(cervical_cancer_data_clean$HISTOLOG, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by MARGINS
plot(prop.table(table(cervical_cancer_data_clean$MARGINS, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by PELLYMPH
plot(prop.table(table(cervical_cancer_data_clean$PELLYMPH, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by MAXDEPTH
plot(prop.table(table(cervical_cancer_data_clean$MAXDEPTH,
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by SIZE
plot(prop.table(table(cervical_cancer_data_clean$SIZE, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])

# Relapse rate by AGE
plot(prop.table(table(cervical_cancer_data_clean$AGE, 
    cervical_cancer_data_clean$RECURRN_Y_or_N), 1)[,2])


```


## Multicollinearity

```{r}
#Checking if any predictor variables are correlated 

# Corr size and age:
cor(cervical_cancer_data_clean$SIZE, cervical_cancer_data_clean$AGE)

# Corr depth and age:
cor(cervical_cancer_data_clean$MAXDEPTH, cervical_cancer_data_clean$AGE)

# Corr depth and size:
cor(cervical_cancer_data_clean$MAXDEPTH, cervical_cancer_data_clean$SIZE)

plot(cervical_cancer_data_clean$MAXDEPTH,
    cervical_cancer_data_clean$SIZE, col="red")

# All correlations are small, multicollinearity does not appear to be a problem.

```

## Models (with no interactions)

```{r}

# logit
cervical_cancer.fit_all_pred_logit <-  
    glm(RECURRN_Y_or_N ~ 
    ADJ_RAD + AGE + CLS +
    GRAD + HISTOLOG + MARGINS + MAXDEPTH + PELLYMPH + SIZE, 
    data=cervical_cancer_data_clean, family=binomial(link = "logit"))

cervical_cancer.fit_all_pred_logit$aic

# probit
cervical_cancer.fit_all_pred_binomial_probit <-  glm(RECURRN_Y_or_N ~ 
    ADJ_RAD + AGE + CLS +
    GRAD + HISTOLOG + MARGINS + MAXDEPTH + PELLYMPH + SIZE, 
    data=cervical_cancer_data_clean, family=binomial(link="probit"))

cervical_cancer.fit_all_pred_binomial_probit$aic

# cloglog
cervical_cancer.fit_all_pred_binomial_cloglog <-  glm(RECURRN_Y_or_N ~ 
    ADJ_RAD + AGE + CLS +
    GRAD + HISTOLOG + MARGINS + MAXDEPTH + PELLYMPH + SIZE, 
    data=cervical_cancer_data_clean, family=binomial(link="cloglog"))

cervical_cancer.fit_all_pred_binomial_cloglog$aic

# Binomial with a probit link has a slightly lower AIC value
# Binomial with a logit link is easier to interpret

# Select binomial logit
cervical_cancer.fit_all_pred <- 
    cervical_cancer.fit_all_pred_logit

summary(cervical_cancer.fit_all_pred)

#Hosmer-Lemeshow test on basic model:
hl <- hoslem.test(cervical_cancer_data_clean$RECURRN_Y_or_N, 
    fitted(cervical_cancer.fit_all_pred))
hl
# Since p-value>0.05, there is no evidence of poor fit

# Importance of each variable
varImp(cervical_cancer.fit_all_pred)

# Visualization of importance of each var
cervical_cancer.fit_all_pred %>% vip()

```

## Confidence Intervals 

```{r}
#Confint function to obtain confidence intervals for the coefficient estimates
confint(cervical_cancer.fit_all_pred)

#Odds Ratios for Model Predictors
exp(cbind(OR = coef(cervical_cancer.fit_all_pred), confint(cervical_cancer.fit_all_pred)))
## THUS CLS, MAXDEPTH associated with increased odds of relapsing
# see: https://rpubs.com/mascha/breastCancerSecond

anova(cervical_cancer.fit_all_pred)
```

## Model Selection (without interactions)

```{r, results="hide"}

# Fit no predictors
cervical_cancer.fit_no_pred <- glm(RECURRN_Y_or_N ~ 1, 
  data=cervical_cancer_data_clean, family=binomial)

# Step backwards through model finding the best fit
# Starting with all predictors
cervical_cancer.fit_no_interaction_back <- 
    step(cervical_cancer.fit_all_pred, direction="backward", test = "Chisq")

print("")
print("Start of forward model selection")
print("")

# Step forwards through model finding the best fit
# Starting with no predictors
cervical_cancer.fit_no_interaction_forward <- 
    step(cervical_cancer.fit_no_pred, scope =~ ADJ_RAD + AGE + CLS + GRAD + 
    HISTOLOG + MARGINS + MAXDEPTH + PELLYMPH + SIZE, 
    direction="forward", test = "Chisq")

print("")
print("Start of both directions model selection")
print("")

# Step both directions through model finding the best fit
# Starting with all predictors
cervical_cancer.fit_no_interaction_both = step(cervical_cancer.fit_all_pred, direction = "both", test="Chisq")


```

## Model comparison (without interactions)

```{r, results="hide"}

summary(cervical_cancer.fit_no_interaction_back)
summary(cervical_cancer.fit_no_interaction_forward)
summary(cervical_cancer.fit_no_interaction_both)

# All selected models without interactions are the same

```


## Model selected (without interactions)

```{r}

# AIC: 267.09
# Backward, forward and both selected model are the same
cervical_cancer.fit_no_interaction <- cervical_cancer.fit_no_interaction_both

summary(cervical_cancer.fit_no_interaction)

# Pearson standard residuals
plot(cervical_cancer.fit_no_interaction, which=1)

```


## Model selection with 2nd degree interactions

```{r}

# Model with 2nd degree interactions
cervical_cancer.fit_all_interactions <- 
    glm(RECURRN_Y_or_N ~ 
    (ADJ_RAD + AGE + CLS +
    GRAD + HISTOLOG + MARGINS + MAXDEPTH + PELLYMPH + SIZE)^2, 
    data=cervical_cancer_data_clean, family=binomial(link = "logit"))

summary(cervical_cancer.fit_all_interactions)

# Importance of each variable
varImp(cervical_cancer.fit_all_interactions)

# Visualization of importance of each var
cervical_cancer.fit_all_interactions %>% vip()

```


## Models with 2nd degree interactions

```{r,  results="hide"}

# Stepping backwards through model doesn't result in a good model and takes far 
# too long

# Step forwards through model finding the best fit
# Starting with no predictors
cervical_cancer.fit_all_interactions_forward <- 
    step(cervical_cancer.fit_no_pred, scope =~ (ADJ_RAD + AGE + CLS + GRAD + 
    MAXDEPTH + PELLYMPH + SIZE)^2, 
    direction="forward", test = "Chisq")

print("")
print("Start of both directions model selection")
print("")

# Step both directions through model finding the best fit
# Starting with all predictors with scope of 2nd degree interactions
cervical_cancer.fit_all_interactions_both <- 
    step(cervical_cancer.fit_all_pred,
    scope =~ (ADJ_RAD + AGE + CLS + GRAD + HISTOLOG +
    MARGINS + MAXDEPTH + PELLYMPH + SIZE)^2, direction = "both", test="Chisq")

```

## Model comparison (with 2nd degree interactions)

```{r}

summary(cervical_cancer.fit_all_interactions_forward)
summary(cervical_cancer.fit_all_interactions_both)

# Add CLS:SIZE to both to make it purely more complex
cervical_cancer.fit_most_complex <- 
    glm(formula = RECURRN_Y_or_N ~ AGE + CLS + MAXDEPTH + PELLYMPH + SIZE + 
    AGE:CLS + PELLYMPH:SIZE + AGE:MAXDEPTH + CLS:SIZE, 
    family = binomial(link = "logit"), data = cervical_cancer_data_clean)

summary(cervical_cancer.fit_most_complex)

anova(cervical_cancer.fit_all_interactions_forward, 
    cervical_cancer.fit_most_complex, test="Chisq" )

# Reject the null that the simpler model sufficiently fits the data as well as 
# the more complex one"

# Check if any interactions can be dropped
drop1(cervical_cancer.fit_most_complex, test="Chisq")

# Drop CLS:SIZE
cervical_cancer.fit_most_complex <- 
    glm(formula = RECURRN_Y_or_N ~ AGE + CLS + MAXDEPTH + PELLYMPH + SIZE + 
    AGE:CLS + PELLYMPH:SIZE + AGE:MAXDEPTH, 
    family = binomial(link = "logit"), data = cervical_cancer_data_clean)

drop1(cervical_cancer.fit_most_complex, test="Chisq")

# Drop AGE:MAXDEPTH
cervical_cancer.fit_most_complex <- 
    glm(formula = RECURRN_Y_or_N ~ AGE + CLS + MAXDEPTH + PELLYMPH + SIZE + 
    AGE:CLS + PELLYMPH:SIZE, 
    family = binomial(link = "logit"), data = cervical_cancer_data_clean)

drop1(cervical_cancer.fit_most_complex, test="Chisq")

# No simpler model would not be rejected for being a sufficient fit

cervical_cancer.fit_selected <- cervical_cancer.fit_most_complex

```


## Add back taken out values after a model is selected

```{r,  results="hide"}

# Start from unclean data again
cervical_cancer_data_clean_final <- cervical_cancer_data_factor

# Only delete NA rows for the columns in the model and patients that 
# didn't have a last follow up
cervical_cancer_data_clean_final <- 
    cervical_cancer_data_clean_final[complete.cases(
    cervical_cancer_data_clean_final[,
    c("AGE", "CLS", "MAXDEPTH", "PELLYMPH", "SIZE")]),]

# More data included
cervical_cancer_data_clean_final

```

# Selected model with all data that can be used

```{r}

cervical_cancer.fit_selected <- 
    glm(formula = RECURRN_Y_or_N ~ AGE + CLS + MAXDEPTH + PELLYMPH + SIZE + 
    AGE:CLS + PELLYMPH:SIZE, 
    family = binomial(link = "logit"), data = cervical_cancer_data_clean_final)

summary(cervical_cancer.fit_selected)

#Hosmer-Lemeshow test on basic model:
hl <- hoslem.test(cervical_cancer_data_clean_final$RECURRN_Y_or_N, 
    fitted(cervical_cancer.fit_selected))
hl
# Since p-value>0.05, there is no evidence of poor fit. 
# p-value is much higher than for the simple model giving more evidence that
# this model is a better fit

# Pearson standard residuals
plot(cervical_cancer.fit_selected, which=1)

```

#Odds Ratios

```{r}

estim <- cbind(summary(cervical_cancer.fit_selected)$coef[,"Estimate"])

#age cls0
exp(estim[2])
#age cls1
exp(estim[2])*exp(estim[7])
#age cls2
exp(estim[2])*exp(estim[8])
#MAXDEPTH
exp(estim[5])
# Size PELLYMPH0
exp(estim[7])
# Size PELLYMPH1
exp(estim[2])*exp(estim[10])

```


# Predicting grouped risk based on odds ratios

```{r}

low_risk <- data.frame("AGE" = 60, "CLS" = as.factor(0), "MAXDEPTH" = 5, "SIZE" = 15, "PELLYMPH"=as.factor(1))

moderate_risk <- data.frame("AGE" = 40, "CLS" = as.factor(1), "MAXDEPTH" = 20, "SIZE" = 5, "PELLYMPH"=as.factor(0))

high_risk <- data.frame("AGE" = 30, "CLS" = as.factor(0), "MAXDEPTH" = 40, "SIZE" = 15, "PELLYMPH"=as.factor(0))

# Low risk
exp(predict(cervical_cancer.fit_selected, low_risk))

# Moderate risk
exp(predict(cervical_cancer.fit_selected, moderate_risk))

# High risk
exp(predict(cervical_cancer.fit_selected, high_risk))

# Baseline relapse rate
sum(cervical_cancer_data_clean_final$RECURRN_Y_or_N)/length(cervical_cancer_data_clean_final$RECURRN_Y_or_N)


```


# Survival analysis

```{r}
data1 <- Surv(as.numeric(cervical_cancer_data_clean$FU_DATE-cervical_cancer_data_clean$SURGDATE))
f1 <- survfit((data1) ~1, data = cervical_cancer_data_clean)
f1
autoplot(f1)

```




## Machine Learning

```{r}

# optimize selected model by the use of repeated cross validation ‘control’ variable - “fitControl”. By that we will be able to limit model overfitting and make sure that it will then generalize well to an unseen dataset

fitControl <- trainControl(method = "repeatedcv",
                           #number of folds is 10 by default
                           repeats = 3, 
                           savePredictions = T)

glmCancerFit <- train(as.factor(RECURRN_Y_or_N) ~ AGE + CLS + 
    MAXDEPTH + PELLYMPH + SIZE + AGE:CLS + PELLYMPH:SIZE,
    data=cervical_cancer_data_clean_final, method="glm", 
    family="binomial", trControl = fitControl)

glmFitAcc <- train(as.factor(RECURRN_Y_or_N) ~ AGE + CLS + 
    MAXDEPTH + PELLYMPH + SIZE + AGE:CLS + PELLYMPH:SIZE, 
    data=cervical_cancer_data_clean_final,
    method="glm", metric="Accuracy", trControl = fitControl) %>% print
### our model is 94% accurate...

# Train/Set Test
set.seed(1234)
create_train_test <- function(data, size = 0.8, train = TRUE) {
    n_row = nrow(data)
    total_row = size * n_row
    train_sample <- 1: total_row
    if (train == TRUE) {
        return (data[train_sample, ])
    } else {
        return (data[-train_sample, ])
    }
}

# split the data between a train set and a test set
data_train <- create_train_test(cervical_cancer_data_clean_final, 0.8, train = TRUE)
data_test <- create_train_test(cervical_cancer_data_clean_final, 0.8, train = FALSE)
dim(data_train)

# Predicted values
predict <- predict(cervical_cancer.fit_selected, data_test, type = 'response')

# confusion matrix to count the number of times True instances are classified are False
table_cancer <- table(data_test$RECURRN_Y_or_N, predict > 0.5)
table_cancer

# Model accuracy calculated by summing the true pos + true neg over the total obvservation
 accuracy_Test <- sum(diag(table_cancer)) / sum(table_cancer)
 accuracy_Test
## 97.6% accuracy - good
 
```

## ROC

```{r}

# ROC-Receiving Operating Characteristic, measures the classifier’s performance using the proportion of positive data points correctly considered as positive (True Positive Rate or Sensitivity) and the proportion of negative data points that are mistakenly considered as positive (False Positive Rate or Fall-out)

preds = predict(cervical_cancer.fit_selected,type = c("response"))
roc1 = roc(RECURRN_Y_or_N ~ preds, data=cervical_cancer_data_clean_final)

# Plot ROC
plot(roc1)

# Area under curve
roc1$auc
# 0.8017 - decent

```

# Plots/Graphs

```{r}
# Age vs Relapse
age_vs_relapse <- 
  cervical_cancer_data_clean_final %>%
  count(AGE,RECURRN_Y_or_N) %>% 
  mutate(AGE=factor(AGE)) %>% 
  ggplot(aes(AGE,n,fill=RECURRN_Y_or_N))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=10, size = 7))+
  labs(subtitle = 'Distribution of Records of Relapse According to Age') +
  geom_vline(aes(xintercept=median(as.numeric(AGE), na.rm = TRUE)), color="red", linetype="dashed", size=1) 
  scale_x_discrete(breaks = seq(0, 100, by = 5))
  
cowplot::plot_grid(age_vs_relapse)

# CLS vs relapse
CLS_vs_relapse <- 
  cervical_cancer_data_clean_final %>%
  count(CLS,RECURRN_Y_or_N) %>% 
  mutate(CLS=factor(CLS)) %>% 
  ggplot(aes(CLS,n,fill=RECURRN_Y_or_N))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=10, size = 7))+
  labs(subtitle = 'Distribution of Records of Relapse According to CLS')

  
cowplot::plot_grid(CLS_vs_relapse)

#Max depth vs relapse
MAXDEPTH_vs_relapse <- 
  cervical_cancer_data_clean_final %>%
  count(MAXDEPTH,RECURRN_Y_or_N) %>% 
  mutate(MAXDEPTH=factor(MAXDEPTH)) %>% 
  ggplot(aes(MAXDEPTH,n,fill=RECURRN_Y_or_N))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=10, size = 7))+
  scale_x_discrete(breaks = seq(0, 30, by = 1))+
  labs(subtitle = 'Distribution of Records of Relapse According to MAXDEPTH')

cowplot::plot_grid(MAXDEPTH_vs_relapse)

#Box plot age vs relapse
age_relapse_box <- ggplot(cervical_cancer_data_clean_final, aes(RECURRN_Y_or_N, AGE)) + geom_boxplot(fill="skyblue") + xlab("Likelihood of Cervical Cancer") + ylab("Patient Age (years)") + geom_point(aes(fill="black"), size = 1, alpha = 0.2, position = position_jitterdodge(), show.legend = FALSE) + theme_bw()

plot(age_relapse_box)

#Box plot maxdepth vs relapse
maxdepth_relapse_box <- ggplot(cervical_cancer_data_clean_final, aes(RECURRN_Y_or_N, MAXDEPTH)) + geom_boxplot(fill="skyblue") + xlab("Likelihood of Cervical Cancer") + ylab("Depth of Tumor") + geom_point(aes(fill="black"), size = 1, alpha = 0.2, position = position_jitterdodge(), show.legend = FALSE) + theme_bw()

plot(maxdepth_relapse_box)

```


```



