---
title: "STA457 Final Project"
author: "Camilly Fuentes"
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Economic indicators that can impact price movements.

-Exchange Rates (USD/GHS & EUR/USD): Cocoa is traded in USD, and fluctuations in exchange rates affect prices.
-Inflation & Interest Rates: Higher inflation and interest rates can influence investor behavior and commodity prices.
-Global GDP Growth: Can help capture trends in global demand.
-Commodity Price Indices: Other agricultural commodities like coffee or sugar may have correlated movements.

```{r}
library(tidyverse)

# Load the cocoa futures price dataset
cocoa_data <- read.csv("/Users/camillyfuentes/Downloads/Daily Prices_ICCO.csv", stringsAsFactors = FALSE)

# Inspect the first few rows
head(cocoa_data)

# Check structure and summary of the data
str(cocoa_data)
summary(cocoa_data)

# Convert DATE column to Date format
cocoa_data$Date <- as.Date(cocoa_data$Date, format="%d/%m/%Y")

# Check for missing values
colSums(is.na(cocoa_data))

# Handle missing values (forward-fill missing prices if applicable)
cocoa_data <- cocoa_data %>% fill(everything(), .direction = "down")

# Remove duplicate rows
cocoa_data <- distinct(cocoa_data)

# Save cleaned cocoa price dataset
write.csv(cocoa_data, "cleaned_Cocoa_Prices.csv", row.names = FALSE)

# Print a message indicating completion
cat("Data cleaning completed. Cleaned data saved as 'cleaned_Ghana_data.csv' and 'cleaned_Cocoa_Prices.csv'\n")

```

```{r}
# Load necessary libraries
library(tidyverse)

# Load the dataset
ghana_data <- read.csv("/Users/camillyfuentes/Downloads/Ghana_data.csv", stringsAsFactors = FALSE)

# Inspect the first few rows
head(ghana_data)

# Check structure and summary of the data
str(ghana_data)
summary(ghana_data)

# Convert DATE column to Date format
ghana_data$DATE <- as.Date(ghana_data$DATE, format="%Y-%m-%d")

# Check for missing values
colSums(is.na(ghana_data))

# Handle missing values:
# - If PRCP is missing, assume 0 (no precipitation)
ghana_data$PRCP[is.na(ghana_data$PRCP)] <- 0

# - If temperature values (TAVG, TMAX, TMIN) are missing, fill with column mean
ghana_data$TAVG[is.na(ghana_data$TAVG)] <- mean(ghana_data$TAVG, na.rm = TRUE)
ghana_data$TMAX[is.na(ghana_data$TMAX)] <- mean(ghana_data$TMAX, na.rm = TRUE)
ghana_data$TMIN[is.na(ghana_data$TMIN)] <- mean(ghana_data$TMIN, na.rm = TRUE)

# Remove duplicate rows
ghana_data <- distinct(ghana_data)

# Save cleaned dataset
write.csv(ghana_data, "cleaned_Ghana_data.csv", row.names = FALSE)

# Print a message indicating completion
cat("Data cleaning completed. Cleaned data saved as 'cleaned_Ghana_data.csv'\n")

```

